---
title: "gcscreen: Complete workflow for 96-well kinetic phenotypic screens"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gcscreen: Complete workflow for 96-well kinetic phenotypic screens}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Overview

`gcscreen` implements a validated pipeline for phenotypic screening using
OD600 kinetic growth curves in 96-well plates. The pipeline comprises:

1. Trapezoidal AUC calculation with baseline correction
2. B-score positional-bias correction (median polish)
3. Normalisation to the plate-run control median
4. Plate quality metrics: CV, Z'-factor, SSMD, robust Z'-factor
5. Per-compound ANOVA and linear mixed model (LMM, plate as random effect)
6. Benjamini-Hochberg FDR, dual-convergence hit calling

## Input data format

Each plate-run is a data frame with columns:

| Column | Type | Description |
|--------|------|-------------|
| `run` | character | Plate-run identifier |
| `plate` | integer | Plate number |
| `well_pos` | integer | Well position 1–96 (row-major) |
| `time` | numeric | Time in hours |
| `od` | numeric | OD600 measurement |
| `Chemical` | character | Compound name or `"CONTROL"` |

A standard 96-well layout has controls in columns 1 and 12 (wells 1, 13, 25,
37, 49, 61, 73, 85, 12, 24, 36, 48, 60, 72, 84, 96) and compounds in columns
2–11 (80 wells per plate-run).

## Quick start: `run_screen()`

The simplest way to use `gcscreen` is the high-level `run_screen()` wrapper.
Because it fits a LMM per compound, it may take a few minutes on a full screen.

```{r quickstart, eval = FALSE}
library(gcscreen)
data(example_screen)

res <- run_screen(
  platerun_list = example_screen,
  time_max      = 60,
  fdr_threshold = 0.05,
  auc_rel_max   = 0.99
)

# Confirmed hits (ANOVA ∩ LMM, FDR < 0.05)
res$hits[, c("Chemical", "AUC_rel", "FDR", "FDR_lmm")]
```

```
#> Chemical     AUC_rel   FDR     FDR_lmm
#> Cycloheximide  0.039  <0.001  <0.001
#> Thimerosal     0.062  <0.001  <0.001
#> Lanoconazole   0.127  <0.001  <0.001
#> ...
```

## Step-by-step usage

### 1. Well layout utilities

```{r layout}
library(gcscreen)

# Map well positions to row/column indices
well_to_rowcol(c(1L, 12L, 13L, 96L))
```

### 2. AUC calculation

```{r auc}
data(example_screen)
run_df <- example_screen[["placa_2_rep_A"]]

# Select one well, sort by time, baseline-correct, compute AUC
w1 <- run_df[run_df$well_pos == 13L, ]   # a control well (col 1, row 2)
w1 <- w1[order(w1$time), ]
od_corr <- w1$od - w1$od[1L]
od_corr[od_corr < 0] <- 0

auc_val <- auc_trapz(w1$time, od_corr)
cat("AUC for well 13:", round(auc_val, 3), "\n")
```

### 3. B-score positional-bias correction

```{r bscore}
# Simulate a plate-run AUC vector with a row gradient
set.seed(42)
wpos    <- 1:96
rc      <- well_to_rowcol(wpos)
auc_raw <- rnorm(96, mean = 40, sd = 2) + rc$row_idx * 1.5

auc_corr <- bscore_plate(wpos, auc_raw)

# Row-AUC correlation in compound wells before and after correction
cmpd_idx <- which(rc$col_idx %in% 2:11)
cat("Pearson r(row, AUC) before B-score:",
    round(cor(rc$row_idx[cmpd_idx], auc_raw[cmpd_idx]),  3), "\n")
cat("Pearson r(row, AUC) after  B-score:",
    round(cor(rc$row_idx[cmpd_idx], auc_corr[cmpd_idx]), 3), "\n")
```

### 4. Plate quality assessment

```{r qc}
# Compute AUC for all control wells of one plate-run
ctrl_df <- run_df[run_df$Chemical == "CONTROL", ]
ctrl_auc <- tapply(seq_len(nrow(ctrl_df)), ctrl_df$well_pos, function(idx) {
  sub <- ctrl_df[idx, ]
  sub <- sub[order(sub$time), ]
  od  <- sub$od - sub$od[1L]; od[od < 0] <- 0
  auc_trapz(sub$time, od)
})

qc <- plate_qc(as.numeric(ctrl_auc), cv_warn = 20)
cat(sprintf("CV: %.1f%%  |  flag: %s\n", qc$cv, qc$cv_flag))
```

### 5. Normalisation

```{r normalize}
# Normalise compound AUC to the control median
ctrl_vec <- as.numeric(ctrl_auc)
cmpd_vec <- c(2.1, 18.4, 35.0, 38.9)   # example values
auc_rel  <- normalize_controls(cmpd_vec, ctrl_vec)
cat("AUC_rel:", round(auc_rel, 3), "\n")
```

### 6. Per-compound statistics (single compound)

```{r stats-single}
# ANOVA: 3 compound replicates vs 48 control replicates (3 plate-runs × 16)
ctrl_all <- rnorm(48, mean = 40, sd = 3)
cmpd_obs <- c(4.2, 3.8, 4.5)            # a strong inhibitor
pval <- screen_anova(cmpd_obs, ctrl_all)
cat("ANOVA p-value:", format(pval, digits = 3), "\n")
```

### 7. Full statistical screen and hit calling

Running `run_anova_screen()` and `run_lmm_screen()` on a full screen is
computationally intensive. The code below is correct but not evaluated in the
vignette to keep build times short.

```{r full-screen, eval = FALSE}
data(example_screen)
res <- run_screen(example_screen)

# Inspect QC table
head(res$qc)

# Top hits by ANOVA FDR
head(res$anova_results)

# Confirmed hits (ANOVA ∩ LMM)
res$hits
```

### 8. Manual hit calling

If you have pre-computed ANOVA and LMM result tables you can call hits
directly:

```{r hit-calling, eval = FALSE}
hits <- call_hits(
  anova_res     = res$anova_results,
  lmm_res       = res$lmm_results,
  fdr_threshold = 0.05,
  auc_rel_max   = 0.99,
  exclude       = "Chicago sky blue 6B"   # optical artefact
)
```

## Interpreting results

| Column | Meaning |
|--------|---------|
| `AUC_rel` | Mean relative AUC (compound / control median). < 1 = growth inhibition. |
| `FDR` | BH-FDR from ANOVA. |
| `FDR_lmm` | BH-FDR from LMM (plate as random effect). Both must be < `fdr_threshold`. |

## Plate layout conventions

The default layout assumes controls in columns 1 and 12 (`ctrl_cols = c(1, 12)`)
and compounds in columns 2–11. Pass a different `ctrl_cols` to `bscore_plate()`
and `run_screen()` to accommodate alternative layouts.

## References

- Zhang J-H, Chung TDY, Oldenburg KR (1999). A Simple Statistical Parameter
  for Use in Evaluation and Validation of High Throughput Screening Assays.
  *J Biomol Screen* 4(2):67–73.
- Zhang XD (2007). A pair of new statistical parameters for quality control in
  RNA interference high-throughput screening assays. *Genomics* 89(4):552–61.
- Bates D et al. (2015). Fitting Linear Mixed-Effects Models Using lme4.
  *J Stat Softw* 67(1):1–48.
